### 2017-12-11 submission/manuscript.Rmd
complete intro more specific sentence structure

Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. Many of the studies, along with the previous meta-analysis, focus on identifying biomarkers or individual microbes but do not critically investigate how the community changes in CRC.

Targeting the identification of biomarkers within stool seems logical since the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free. Stool offers an easy and cost-effective way in which to stratify CRC risk within the population. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to what is actually happening on the adenoma or carcinoma. Some studies have begun to try and assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Since the stool possibly represents an agregate community of the overall gastrointestinal (GI) tract, sampling the adenoma or carinoma directly could more accurately identify important microbes which could be drastically different then what is identified in stool. Sampling the tissue directly may provide clearer answers but it is not without problems. Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa. Additionally, these samples would contain much more host DNA then microbial DNA, potentially limiting the types of analysis that can be done. It has also been well published that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these many differences, one question our meta-analysis aims to answer is whether it is possible to find consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly mouth-associated microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014]. The main bacteria of interest arising from this set of microbes has been those within the *Fusobacterium* genus. Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014]. Based on this discrepency, the second question we want to answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.

The heavy concentration on the identification of biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.  This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area, that have focused predominately on biomarker identification. Since it has not been investigated in detail the final question that this meta-analysis aims to answer is whether there are consistent detectable community differences as disease severity increases.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC). We observed that the community changes as disease severity worsens and that this community is important for disease classification. However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for. Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis, along with other studies, focused on identifying biomarkers or individual microbes but do not critically investigate how the community changes in CRC.

Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to the community on the adenoma or carcinoma. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. One advantage of tissue is that stool possibly represents an agregate community of the overall gastrointestinal (GI) tract and sampling the adenoma or carinoma directly could more accurately identify important microbes for CRC. Sampling the tissue directly may provide clearer answers but is not without limitations. Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these many differences that could arise between stool and tissue, one question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

This intense focus on identifying biomarkers has found strong associations with resident mouth microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014]. The main bacteria of interest, arising from this set of microbes, has been those within the *Fusobacterium* genus. Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014]. Based on this discrepency, the second question we want to answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.

The identification of microbial biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.  This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area to date, that focus predominately on biomarker identification. Yet, how these CRC-associated microbes interact with their community may be an important component to consider, and has not been investigated in great detail by other meta-analyses. In response to this gap, our study aims to answer whether there are consistent detectable community differences as disease severity increases.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC). Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification. However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for. Although we analyzed data sets that sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

### 2017-12-11 submission/manuscript.tex
complete intro more specific sentence structure

Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in polyp formation using mouse models of CRC. There have
also been numerous case/control studies investigating the microbiota in
the formation of both adenoma and carcinoma. Recently, a meta-analysis
was published investigating whether specific biomarkers could be
consistently identified using multiple data sets {[}11{]}. Many of the
studies, along with the previous meta-analysis, focus on identifying
biomarkers or individual microbes but do not critically investigate how
the community changes in CRC.

Targeting the identification of biomarkers within stool seems logical
since the current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Stool offers an easy and
cost-effective way in which to stratify CRC risk within the population.
Although stool represents an easy and less invasive way to assess risk,
it is not clear how reflective this sample is to what is actually
happening on the adenoma or carcinoma. Some studies have begun to try
and assess this in health and disease but are limited by their sample
size {[}12,13{]}. Since the stool possibly represents an agregate
community of the overall gastrointestinal (GI) tract, sampling the
adenoma or carinoma directly could more accurately identify important
microbes which could be drastically different then what is identified in
stool. Sampling the tissue directly may provide clearer answers but it
is not without problems. Due to bowel prep the communities left for
sampling may not be reflective of the resident microbiota, but rather a
collection of what is able to keep adhered to the mucosa. Additionally,
these samples would contain much more host DNA then microbial DNA,
potentially limiting the types of analysis that can be done. It has also
been well published that low biomass samples can be very difficult to
work with and results can end up being study dependent due to the
randomness of contamination {[}14{]}. Due to these many differences, one
question our meta-analysis aims to answer is whether it is possible to
find consistent patterns that emerge across studies regardless of
whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly
mouth-associated microbes as potential CRC-associated microbes in the GI
tract {[}15--17{]}. The main bacteria of interest arising from this set
of microbes has been those within the \emph{Fusobacterium} genus. Yet,
the question remains as to whether or not this is indeed the most
important genera to be focusing on, since many microbiota-based studies
typically have identified a collection of oral microbes rather than
single species from a single genera {[}16,17{]}. Based on this
discrepency, the second question we want to answer with this
meta-analysis is if there is one dominant CRC-associated genera that can
be identified across studies.

The heavy concentration on the identification of biomarkers has mostly
dominated the study of CRC and has had an unattended consequence of
reducing the focus on changes that occur within the underlying resident
community. This has been borne out by the majority of previous studies,
within stool, tissue, and the only meta-analysis investigating this
area, that have focused predominately on biomarker identification. Since
it has not been investigated in detail the final question that this
meta-analysis aims to answer is whether there are consistent detectable
community differences as disease severity increases.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether the diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Next, using Random
Forest models, we analyzed whether the full community or only the
CRC-associated genera resulted in better model classification based on
the area under the curve (AUC). We observed that the community changes
as disease severity worsens and that this community is important for
disease classification. However, since the changes in community were
subtle for adenoma we also examined what effect and sample size the
studies that were used were adequately powered for. Although we analyzed
data sets which sampled large numbers of individuals, our results
indicate the individual studies were underpowered for detecting effect
size differences of 10\% or below between the case and control groups.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis, along
with other studies, focused on identifying biomarkers or individual
microbes but do not critically investigate how the community changes in
CRC.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}12,13{]}. One advantage of tissue is that stool
possibly represents an agregate community of the overall
gastrointestinal (GI) tract and sampling the adenoma or carinoma
directly could more accurately identify important microbes for CRC.
Sampling the tissue directly may provide clearer answers but is not
without limitations. Due to bowel prep the communities left for sampling
may not be reflective of the resident microbiota, but rather a
collection of what is able to keep adhered to the mucosa. Additionally,
these samples contain more host DNA, potentially limiting the types of
analysis that can be done. It is well known that low biomass samples can
be very difficult to work with and results can end up being study
dependent due to the randomness of contamination {[}14{]}. Due to these
many differences that could arise between stool and tissue, one question
our meta-analysis aims to answer is whether there are consistent
patterns that emerge across studies regardless of whether they used
stool or tissue samples.

This intense focus on identifying biomarkers has found strong
associations with resident mouth microbes as potential CRC-associated
microbes in the GI tract {[}15--17{]}. The main bacteria of interest,
arising from this set of microbes, has been those within the
\emph{Fusobacterium} genus. Yet, the question remains as to whether or
not this is indeed the most important genera to be focusing on, since
many microbiota-based studies typically have identified a collection of
oral microbes rather than single species from a single genera
{[}16,17{]}. Based on this discrepency, the second question we want to
answer with this meta-analysis is if there is one dominant
CRC-associated genera that can be identified across studies.

The identification of microbial biomarkers has mostly dominated the
study of CRC and has had an unattended consequence of reducing the focus
on changes that occur within the underlying resident community. This has
been borne out by the majority of previous studies, within stool,
tissue, and the only meta-analysis investigating this area to date, that
focus predominately on biomarker identification. Yet, how these
CRC-associated microbes interact with their community may be an
important component to consider, and has not been investigated in great
detail by other meta-analyses. In response to this gap, our study aims
to answer whether there are consistent detectable community differences
as disease severity increases.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether diversity changes throughout
disease (control to adenoma to carcinoma) and if it results in an
increased relative risk (RR) for adenoma or carcinoma. We then assessed
how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Next, using Random
Forest models, we analyzed whether the full community or only the
CRC-associated genera resulted in better model classification based on
the area under the curve (AUC). Our results suggest that the community
changes as disease severity worsens and that this community is important
for disease classification. However, since the changes in community were
subtle for adenoma we also examined what effect and sample size the
studies that were used were adequately powered for. Although we analyzed
data sets that sampled large numbers of individuals, our results
indicate the individual studies were underpowered for detecting effect
size differences of 10\% or below between the case and control groups.
